Scaffolder - Software for Reproducible Genome Scaffolding.

### 2011-03-02 src/article.tex
Proof read introduction

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than 1,000 nucleotides
in length. Therefore to resolve a sequence longer than this, such as a complete
genome, these numerous smaller fragments must be pieced together. The process
of joining reads into longer sequences is the `assembly' stage of a genome
project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle.
Each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough or multiple overlaps between sequence reads and is analogous to missing
pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has led
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of the many available individual reads, which represent
the assembly of reads into the longest possible sequences given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into larger sequence lengths. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of required steps in the scaffolding process is outlined
below.

@@

The sequencing process generates reads from either strand of the DNA helix, the
resulting contigs constructed from these reads may point in either direction of
the DNA. Orientating contigs to point in the same direction and represent the
same strand requires producing the reverse complement of sequences where
necessary. In the case of microbial genomes this orientation may be to the
5'-3' direction leading away from the origin of replication.

@@

Contig ordering determines the placement of observed contigs to best represent
their order in the true genome sequence. The Correct placement of each contig
also highlights any extra-genomic DNA, such as plasmids which are scaffolded
separately from the genomic sequence. The contig order is commonly started at
the contig containing the origin of replication. All subsequent contigs are
then ordered in the 5' - 3' direction of DNA replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
still to be resolved area in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets can be used to infer
contig placement and to build the scaffold
\cite{richter2007,zhao2008,assefa2009}. Recombination between the two genomes
can, however, reduce the effectiveness of this analysis. Repeat regions may be
responsible for many of the gaps in building a genome sequence. The existence
of tandemly repeated nucleotide regions produces many similar reads in the
sequence process. The assembly of reads generated from repeat regions can
result in the assemble collapsing into a single repeat instance or ignored by
conservative assembly algorithms. These regions may be resolved by using
reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use unassembled
sequence reads matching the regions around a scaffold gap to construct
a uniquely overlapping set of reads across the region \cite{tsai2010}. \pb

@@

These methods provide a wide array of approaches for merging contigs into
larger a continuous scaffolds. The scaffolding process may still however
require inserting additional sequence or joining contigs using PCR-derived
information from the laboratory. This manual stage in the finishing process
will moving and editing large blocks of nucleotide text by hand which is subject
to human error-prone and precludes any reproducibility for the steps taken. \pb

The software outlined here, ``Scaffolder'', aims to solve address these
problems by creating a file syntax and software framework for editing a genome
scaffold. Scaffolder uses a specific file format to define how contigs are
joined, additional sequences inserted, and for the specification of unknown
regions. This syntax allows a scaffold to be updated by simply editing the
scaffold file facilitating a reproducible scaffolding process and providing
a concise overview of how the scaffold was constructed. \pb

>>>>

High-throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing,
high-throughput sequencing is limited to producing reads less than 1,000
nucleotides in length. Therefore to resolve a sequence longer than this, such
as a complete genome, these numerous smaller fragments must be pieced together.
The process of joining reads into longer sequences is the `assembly' stage of
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough, or multiple different overlaps between reads and is analogous to
missing pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has led
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases the algorithm
will generate several large assembled regions of sequence (`contigs') composed
of the many available individual reads, which represent the assembly of reads
into the longest possible sequences given the data. Nonetheless, these contigs
represent a fragmented picture of the genome and require additional work to
join together into larger lengths. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of the required steps in the scaffolding process is outlined
below.

@@

The sequencing process generates reads from either strand of the DNA helix the
resulting contigs constructed from these reads may point in either direction of
the DNA. Orientating contigs to point in the same direction and represent the
same strand requires reverse complementing sequences where necessary. In the
case of microbial genomes this orientation may be to the 5' $\rightarrow$ 3'
direction leading away from the origin of replication.

@@

Contig ordering determines the placement of observed contigs to best represent
their order in the true genome sequence. The correct placement of each contig
also highlights any extra-genomic DNA, such as plasmids which are scaffolded
separately from the genomic sequence. The order is commonly started at the
contig containing the origin of replication. All subsequent contigs are then
ordered in the 5' $\rightarrow$ 3' direction of DNA replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
still to be resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is then used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets can be used to infer
contig placement and build the contigs into a scaffold
\cite{richter2007,zhao2008,assefa2009}. Recombination between the two genomes
can, however, reduce the effectiveness of this analysis. Repeat regions may
also be responsible for many of the gaps in building a genome sequence. The
existence of tandemly repeated nucleotide regions produces many similar reads
fromgenome sequencing. The assembly of reads generated from repeat regions can
result in the assembly collapsing into a single repeat instance or ignored by
more conservative assembly algorithms. These regions may however be resolved by
using reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use unassembled
sequence reads matching the regions around a scaffold gap to construct
a uniquely overlapping set of reads across the gap \cite{tsai2010}. \pb

@@

These methods provide a wide array of approaches for merging contigs into
larger continuous scaffold sequences. The scaffolding process may still however
require inserting additional sequence or joining contigs using PCR-derived
information from the laboratory. This manual stage in the finishing process
comprises moving and editing large blocks of nucleotide text by hand which is
subject to human error and precludes any reproducibility of the steps taken.
\pb

The software outlined here, ``Scaffolder'', aims to address these problems by
creating a file syntax and software framework for editing a genome scaffold.
Scaffolder uses a specific file format to define how contigs are joined,
additional sequences inserted, and for the specification of unknown regions.
This syntax allows a scaffold to be updated by simply editing the scaffold
file. Scaffolder facilites a reproducible scaffolding process and provides
a concise overview of how the scaffold was constructed. \pb
